---
title: "Hematopoiesis_2016_demo"
output:
  html_document:
    df_print: paged
---

##### Zhang Yi <synapse@pku.edu.cn>
##### 20220308

In this vignette we will show you a brief guide how to run EpiTrace analysis on a demo dataset from `Corces 2016`. These demo data are from FACS-sorted pure hematopoietic cell populations and sequenced by bulk ATAC-seq. Similar code works with single cell dataset.    

**Reference**:   
Corces MR, Buenrostro JD, Wu B, Greenside PG, Chan SM, Koenig JL, Snyder MP, Pritchard JK, Kundaje A, Greenleaf WJ, Majeti R, Chang HY. Lineage-specific and single-cell chromatin accessibility charts human hematopoiesis and leukemia evolution. Nat Genet. 2016 Oct;48(10):1193-203. doi: 10.1038/ng.3646. Epub 2016 Aug 15. PMID: 27526324; PMCID: PMC5042844.  



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


##### Setting up the environment


```{r setup}
# Not run
# library(devtools)
# devtools::install_github('eulertx/EpiTrace') 

library(dplyr)
library(tidyr)
library(readr)
library(GenomicRanges)
library(reshape2)
library(openxlsx)
library(ggplot2)
library(Matrix)
library(EpiTrace)
library(Seurat)
library(SeuratObject)
library(ggtree)
library(EnsDb.Hsapiens.v86)
library(patchwork)
library(ArchR)

```

##### Loading demonstration data

```{r load data}
data("clock_gr_list")
data("hematopoiesis.2018")
```

Generally, `clock_gr_list` is a list of GRanges each corresponds to a set of clock-like DML (clockDML) in your favourite organism. Here we provide human data in the package. You can make your own list for other organisms.  
There are three sets of human clockDML we provides in `clock_gr_list`:  
    *Mitosis*: 616 clock-like DML compiled from papers of `Yang et al 2016`, `Teschendorff 2020`, and `Youn and Wang 2018`.  
    *Chronology*: 125625 clock-like DML/DMR computed from genome-wide bisulfite capture sequencing data and a compilation of 450k methyl-array data, removing the ones that overlaps with *Mitosis* dataset.   
    *solo_WCGW*: 1669720 solitary *WCGW* motif in partial methylated domain, from `Zhou et al 2018`. 
    
**Reference**:   
Zhou W, Dinh HQ, Ramjan Z, Weisenberger DJ, Nicolet CM, Shen H, Laird PW, Berman BP. DNA methylation loss in late-replicating domains is linked to mitotic cell division. Nat Genet. 2018 Apr;50(4):591-602. doi: 10.1038/s41588-018-0073-4. Epub 2018 Apr 2. PMID: 29610480; PMCID: PMC5893360.  
Yang Z, Wong A, Kuh D, Paul DS, Rakyan VK, Leslie RD, Zheng SC, Widschwendter M, Beck S, Teschendorff AE. Correlation of an epigenetic mitotic clock with cancer risk. Genome Biol. 2016 Oct 3;17(1):205. doi: 10.1186/s13059-016-1064-3. PMID: 27716309; PMCID: PMC5046977.  
Teschendorff AE. A comparison of epigenetic mitotic-like clocks for cancer risk prediction. Genome Med. 2020 Jun 24;12(1):56. doi: 10.1186/s13073-020-00752-3. PMID: 32580750; PMCID: PMC7315560.  
Youn A, Wang S. The MiAge Calculator: a DNA methylation-based mitotic age calculator of human tissue types. Epigenetics. 2018;13(2):192-206. doi: 10.1080/15592294.2017.1389361. Epub 2018 Feb 6. PMID: 29160179; PMCID: PMC5873367.  
Horvath S. DNA methylation age of human tissues and cell types. Genome Biol. 2013;14(10):R115. doi: 10.1186/gb-2013-14-10-r115. Erratum in: Genome Biol. 2015;16:96. PMID: 24138928; PMCID: PMC4015143.  


```{r 0.1}
names(clock_gr_list)
head(clock_gr_list[['Mitosis']])
```

##### Step 1. Initialize peak sets

Peak sets are the annotation of ATAC peaks in your input dataset. Either a dataframe with columns `chr`,`start`,`end` or a GRanges object is fine. Here we provide you with an example GRanges. Initaliztion of peak sets is necessary for following steps.   

```{r initiate peak sets}
initiated_peaks <- Init_Peakset(inputpeak) 
head(initiated_peaks)
```

##### Step 2. Prepare EpiTrace object 

Use `EpiTrace_prepare_object` to prepare the `Seurat` object which is built by `Signac` for following usage. Basically we need two input data at minimal: the peakset and the input count matrix.   

Input count matrix, a row(peaks)_x_col(cells) matrix of reads, could be re-generated from ArchR or Signac.   

Please note that reference genome have to be set here. Currently we have support `hg19` and `hg38` because we have only provided human clockDML datasets. If you use a non-human dataset, you have to set the `non_standard_clock` option to `TRUE` and give your own `clock_gr_list`, which is a list of GRanges each corresponds to a set of clockDML in your favourite organism.   

**NOTE** consistency between ref_genome, initiated_peaks, and the given clockDML is essential.  

This results in a Seurat object of your input matrix and peaks.  

```{r prepare epitrace object}
epitrace_obj <- EpiTrace_prepare_object(initiated_peaks,inputmatrix,cell_metadata$celltype,ref_genome = 'hg19',non_standard_clock = F,clock_gr_list = clock_gr_list)
epitrace_obj
```
##### Step 3. Calculate EpiTrace Age  

We pass the prepared epitrace object to estimate age using `RunEpiTraceAge` command.   

This would add several columns into the meta.data slot of the resulting Seurat object, corresponding to cell age determined by EpiTrace. Generally, early-generated cells are with longer (larger) age estimate, and late-generated cells are with shorter (smaller) age estimate.  

We can see that chromatin accessibility on mitotic clock is most correlated to cell division (age) whilst the other two (chronology clock) are associated with not only mitosis number but also chronological time of a cell lived.    

```{r estimate epitrace age}
library(ggplot2)
epitrace_obj_age_estimated <- RunEpiTraceAge(epitrace_obj)
ggplot(epitrace_obj_age_estimated@meta.data,aes(x=EpiTraceAge_MitosisClock,y=celltype)) + geom_boxplot() + theme_classic() |
  ggplot(epitrace_obj_age_estimated@meta.data,aes(x=EpiTraceAge_ChronologyClock,y=celltype)) + geom_boxplot() + theme_classic() |
  ggplot(epitrace_obj_age_estimated@meta.data,aes(x=EpiTraceAge_AllClock,y=celltype)) + geom_boxplot() + theme_classic() 
```
##### Step 4. Compute the phylogenetic tree based on EpiTrace  

We pass the prepared epitrace object to estimate per-cell-cluster phylogenetic tree using `RunEpiTracePhylogeny` command.   

This generates a list which contains the `assay` (id of assay), `tree` (phylogenetic tree of the clusters), `tree_plot` (a ggtree object).   

If you would like to use other clustering of cells (say, single cell) you have to change the `Idents` of the object. Currently we tend to suggest use well annotated single cell clusters (or in this demo case, known FACS-sorted cell types). 

```{r estimate phylogenetic tree}
phylotree_res <- RunEpiTracePhylogeny(epitrace_obj_age_estimated)
```

We can root the phylogenetic tree with known start (in this case, `HSC`) and re-draw the tree.  

```{r new root}

mitosis_tree <- phylotree_res[['MitosisClock']][[2]]
mitosis_tree <- ape::root(mitosis_tree,outgroup='HSC')
plot(mitosis_tree)

```

We can, of course, run a sub-analysis on a lineage of cells.  

By default this is re-doing a clustering round but you can always remove it by option.  


```{r estimate phylogenetic tree sublineag}
phylotree_res_myeloid <- RunEpiTracePhylogeny(subset(epitrace_obj_age_estimated,celltype %in% c('HSC','MPP','CMP','GMP','Monocyte')))
chronology_tree_myeloid <- phylotree_res_myeloid[['ChronologyClock']][[2]]
chronology_tree_myeloid <- ape::root(chronology_tree_myeloid,outgroup='HSC')
plot(chronology_tree_myeloid)
```
##### Step 5. Compute association between regulatory region accessibility and cell age.  

We pass the age-estimated object to `AssociationOfPeaksToAge` function to compute association result.  

Note that association is most meaningful for a sub-population.  

```{r perform association test}
associated_res_HSC <- AssociationOfPeaksToAge(subset(epitrace_obj_age_estimated,celltype %in% 'HSC'),epitrace_age_name = "EpiTraceAge_MitosisClock")
head(associated_res_HSC)
```


Age-associated regulatory regions are regions whose accessibility is strongly correlated with sample/cell age. We annotate the regions and make an example of finding HMGA2, IL1R1 and ZIC3, which were all reported to be associated with long-aged HSC compartment, and NOS3 -- which is reported to be expressed in HSC and regulate early-stage formation of HSC by activating blood flow -- as well as RUNX1, the known frequent AML translocation partner, associated with younger HSC.  


**Reference**:   
Kumar P, Beck D, Galeev R, Thoms JAI, Talkhoncheh MS, de Jong I, Unnikrishnan A, Baudet A, Subramaniam A, Pimanda JE, Larsson J. HMGA2 promotes long-term engraftment and myeloerythroid differentiation of human hematopoietic stem and progenitor cells. Blood Adv. 2019 Feb 26;3(4):681-691. doi: 10.1182/bloodadvances.2018023986. PMID: 30808686; PMCID: PMC6391669.  
Pietras EM, Mirantes-Barbeito C, Fong S, Loeffler D, Kovtonyuk LV, Zhang S, Lakshminarasimhan R, Chin CP, Techner JM, Will B, Nerlov C, Steidl U, Manz MG, Schroeder T, Passegué E. Chronic interleukin-1 exposure drives haematopoietic stem cells towards precocious myeloid differentiation at the expense of self-renewal. Nat Cell Biol. 2016 Jun;18(6):607-18. doi: 10.1038/ncb3346. Epub 2016 Apr 25. PMID: 27111842; PMCID: PMC4884136.  
North TE, Goessling W, Peeters M, Li P, Ceol C, Lord AM, Weber GJ, Harris J, Cutting CC, Huang P, Dzierzak E, Zon LI. Hematopoietic stem cell development is dependent on blood flow. Cell. 2009 May 15;137(4):736-48. doi: 10.1016/j.cell.2009.04.023. PMID: 19450519; PMCID: PMC2722870.  
Speck NA, Gilliland DG. Core-binding factors in haematopoiesis and leukaemia. Nat Rev Cancer. 2002 Jul;2(7):502-13. doi: 10.1038/nrc840. PMID: 12094236.  



```{r annotation of regulatory regions to genes}

library(ChIPseeker)
library(tradeSeq)
library(ggplot2)
library(ggtree)
library(clusterProfiler)
library(dplyr)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(org.Hs.eg.db)
library(tidyr)
library(reshape2)

txdb = TxDb.Hsapiens.UCSC.hg19.knownGene
associated_res_HSC <- separate(associated_res_HSC,col='locus',into=c('chr','start','end'),remove=F,convert=T)
associated_res_HSC_gr <- makeGRangesFromDataFrame(associated_res_HSC)

findOverlaps(associated_res_HSC_gr,plyranges::reduce_ranges(c(clock_gr_list[[1]],clock_gr_list[[2]])))@from %>% unique -> peaks_overlap_with_clock
associated_res_HSC_gr$locus_type <- 'Others'
associated_res_HSC_gr$locus_type[peaks_overlap_with_clock] <- 'Clock-like DML'


annotatePeak(associated_res_HSC_gr, tssRegion=c(-2000, 500), TxDb=txdb,addFlankGeneInfo=F, flankDistance=50000,annoDb = "org.Hs.eg.db") -> associated_res_HSC_gr_anno 

as.data.frame(associated_res_HSC_gr_anno@anno) -> associated_res_HSC_gr_anno_df
cbind(associated_res_HSC,associated_res_HSC_gr_anno_df %>% dplyr::select(SYMBOL,distanceToTSS,annotation,locus_type) ) -> associated_res_HSC
associated_res_HSC$promoter_scaled_cor <- NA
associated_res_HSC$promoter_scaled_cor[grepl('Promoter',associated_res_HSC$annotation)] <- scale(associated_res_HSC$correlation_of_EpiTraceAge[grepl('Promoter',associated_res_HSC$annotation)])
associated_res_HSC[associated_res_HSC$promoter_scaled_cor>2.1 & !is.na(associated_res_HSC$promoter_scaled_cor) & associated_res_HSC$locus_type %in% 'Others',] %>% arrange(-promoter_scaled_cor) %>% dplyr::select(promoter_scaled_cor,distanceToTSS,SYMBOL,locus_type)

associated_res_HSC[associated_res_HSC$promoter_scaled_cor < -2 & !is.na(associated_res_HSC$promoter_scaled_cor),] %>% arrange(promoter_scaled_cor) %>% dplyr::select(promoter_scaled_cor,distanceToTSS,SYMBOL,locus_type)

```


